Abstraction and Invariance for Algebraically Indexed Types

### 2012-10-23 paper/section-instantiations.tex
Continued work on general isomorphism

\begin{example}[Geometry: change of origin]\label{ex:simpletrans}
We prove the isomorphism
$\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}) \cong
\tyPrim{vec}{e,0}$ where $e$ is an expression of sort $\SynGL{2}$ not containing $t$.
Let 
\begin{align*}
X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}\\
Y & \isDefinedAs\tyPrim{vec}{e,0}\\
I & \isDefinedAs\lambda f\mathord:X.f_0\,0_0\\
J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{e,t}.w +_{e,t,0} v
\end{align*}
Unfolding definitions gives that $\tmSem{I(J(v))}(v,\opsEnv)$ = $0+v$
for any $v\in\mathbb R ^ 2$.  Because $0$ is the identity for vector
addition it follows that $\ctxteq{\Delta}{v\mathord:Y}{I(J(v))}{v}Y$.
To show $\ctxteq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
\autoref{thm:soundness} and reason using the relational semantics.  It
suffices to show $\semeq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$; that
is, given $\rho\in\semSort\Delta$, for any $f$ and $f'$ with
$(f,f')\in\rsem X\rho$, it is the case that 
$(\tmSem{I(J(f))}(f,\opsEnv), f')\in\rsem X\rho$.
Expanding the premise, and setting $B = \semIndexExp{e}\rho$, we have
\begin{eqnarray}\label{eqn:transpremise}
\forall t\in \mathbb R ^ 2,
\forall v\in \mathbb R ^ 2,
f'(Bv + t) = B(f(v)) + t
\end{eqnarray}
Expanding the conclusion, we have to show that for any
$t'\in\mathbb R^2$ and $v'\in\mathbb R^2$, it is the case that
$f'(Bv' + t') = B(v' + f(0)) + t'$.
By instantiating \autoref{eqn:transpremise} thrice, with~$v=t=0$, with~$v=v'$ and~$t=t'$,
and with (crucially)~$v=v'$ and~$t=-Bv'$, and applying a little algebra, we obtain
the desired result.
\qed
\end{example}

@@

\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of this isomorphism; another instance is the
type of vector addition:
\[
\begin{array}{l}
\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1+t_2}\\
\cong\quad
\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}\\
\cong\quad
\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0}
\end{array}
\]
\noindent
\newcommand{\transup}[3]{\mathord\uparrow_{#1}^{#2\mathord:#3}}
\newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}
The maps associated with the isomorphism are 
$I = \lambda f. f_0\ 0_0$ and $J = \lambda y. \Lambda t.\lambda v. \transup A t v (y)$,
%\begin{eqnarray*}
%I = \lambda x. x\ [0]\ 0 & \text{ and } &
%J = \lambda y. \Lambda t.\ \lambda v. \trans A v t(y)
%\end{eqnarray*}
where terms $\transdn A t v : A\to A[0/t]$ and $\transup A t v : A[0/t] \to A$ are defined by induction over the syntax of $A$ (cases omitted for $\transup A t v$ have definition symmetric to $\transdn A t v$):
\begin{align*}
\transdn{\tyPrimNm{unit}} t v &= \lambda x. x \qquad
\transdn{A_1\times A_2} t v = \lambda p. (\transdn{A_1} t v(\pi_1 p),\transdn{A_2}t v(\pi_2 p))\\ 
\transdn{A_1 + A_2} t v &= \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transdn {A_2} t v y)\\ 
\transdn{A_1\to A_2} t v &= \lambda f. \lambda x. \transdn{A_2} t v (f(\transup {A_1} t v x))\\ 
\transdn{\tyPrim{vec}{e}} t v &= \lambda x. x - kv \qquad
\transup{\tyPrim{vec}{e}} t v = \lambda x. x + kv\\ 
\text{ if }e & \equiv k t + e' \text{ for
  some }k\in\setOfIntegers \text{ and }e'\text{ with }t\notin e'
\end{align*}
To show $I(J(y))$ is contextually equivalent to $y$ is an easy
induction on $A$ using the fact that the vector $0$
is an identity for vector addition. To show $J(I(x))$ is contextually
equivalent to $x$ we must apply the abstraction theorem and use
the fact that $\tmSem{\transup A t v}$ and $\tmSem{\transdn A t v}$ 
considered as relations are simply instances of $\rsem A{}{}$.
%\[
%$\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\transup A (x)}) \mid x \in
%\tySem A\}$.
%\]

\paragraph{Scalings}
For the group $\GL{1}$ of scalings we can treat
a real-valued argument to a function as a unit-of-measure with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A
& \cong &
A[1/s]\times \forall s\mathord:\SynGL{1}.A
\end{eqnarray*}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ \quad\cong\quad \tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s^2})\\ \quad\cong\quad
\tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
\end{eqnarray*}
%where %the isomorphism 
The abstraction theorem is used to prove
$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong 1$.
% is proved using
%the abstraction theorem.

%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}

>>>>

\begin{example}[Geometry: change of origin]\label{ex:simpletrans}
We prove the isomorphism
$\Delta\vdash(\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}) \cong
\tyPrim{vec}{e,0}$ where $e$ is an expression of sort $\SynGL{2}$ not containing $t$.
Let 
\begin{align*}
X & \isDefinedAs\forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to\tyPrim{vec}{e,t}
&
Y & \isDefinedAs\tyPrim{vec}{e,0}\\
I & \isDefinedAs\lambda f\mathord:X.f\,[0]\,(0\,[e])\\
J & \isDefinedAs\lambda v\mathord:Y.\Lambda t\mathord:\SynTransl{2}.\lambda w\mathord:\tyPrim{vec}{e,t}.v +[e,0,t] w
\end{align*}
Unfolding definitions gives that $\tmSem{I(J(v))}(v,\opsEnv)$ = $v+0$
for any $v\in\mathbb R ^ 2$.  Because $0$ is the identity for vector
addition it follows that $\ctxteq{\Delta}{v\mathord:Y}{I(J(v))}{v}Y$.
To show $\ctxteq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$ we appeal to
\autoref{thm:soundness} and reason using the relational semantics.  It
suffices to show $\semeq{\Delta}{f\mathord:X}{J(I(f))}{f}{X}$; that
is, given $\rho\in\semSort\Delta$, for any $f$ and $f'$ with
$(f,f')\in\rsem X\rho$, it is the case that 
$(\tmSem{J(I(f))}(f,\opsEnv), f')\in\rsem X\rho$.
Expanding the premise, and setting $B = \semIndexExp{e}\rho$, we have
\begin{eqnarray}\label{eqn:transpremise}
\forall t\in \mathbb R ^ 2,
\forall v\in \mathbb R ^ 2,
f'(Bv + t) = B(f(v)) + t
\end{eqnarray}
Expanding the conclusion, we have to show that for any
$t'\in\mathbb R^2$ and $v'\in\mathbb R^2$, it is the case that
$f'(Bv' + t') = B(v' + f(0)) + t'$.
By instantiating \autoref{eqn:transpremise} thrice, with~$v=t=0$, with~$v=v'$ and~$t=t'$,
and with (crucially)~$v=v'$ and~$t=-Bv'$, and applying a little algebra, we obtain
the desired result.
\qed
\end{example}

@@

\exref{ex:type-iso} and \exref{ex:simpletrans} are special cases of this isomorphism; another instance is the
type of vector addition:
\[
\begin{array}{l}
\forall B\mathord:\SynGL{2}, t_1
%\mathord:\SynTransl{2}
, t_2\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_2}\to\tyPrim{vec}{B,t_1\mathord+t_2}\\
\cong\quad
\forall B\mathord:\SynGL{2}, t_1\mathord:\SynTransl{2}.\tyPrim{vec}{B,t_1}\to\tyPrim{vec}{B,t_1}\\
\cong\quad
\forall B\mathord:\SynGL{2}.\tyPrim{vec}{B,0}
\end{array}
\]
\noindent
\newcommand{\transup}[3]{\mathord\uparrow_{#1}^{#2\mathord:#3}}
\newcommand{\transdn}[3]{\mathord\downarrow_{#1}^{#2\mathord:#3}}

\begin{example}
Let 
\begin{align*}
X & \isDefinedAs \forall t\mathord:\SynTransl{2}.\tyPrim{vec}{e,t}\to A
& 
Y & \isDefinedAs A[0/t]\\
I & \isDefinedAs \lambda f\mathord: X. f\ [0]\ (0\ [e])
& 
J & \isDefinedAs \lambda y\mathord: Y. \Lambda t.\lambda v\mathord:\tyPrim{vec}{e,t}. \transup A t v (y)
\end{align*}
where terms 
\begin{align*}
&
\Delta,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{e,t}\vdash \transup A t v : A[0/t] \to A
\\\text{and } &
\Delta,t\mathord:\SynTransl{2};v\mathord:\tyPrim{vec}{e,t}\vdash \transdn A t v : A\to A[0/t]
\end{align*}
are defined by induction over the syntax of $A$. For brevity we omit types on binders
and note that the omitted cases for $\transup A t v$ 
have definition symmetric to $\transdn A t v$:
\begin{align*}
\transdn{\tyPrimNm{unit}} t v &= \lambda x. x
\qquad 
\transdn{\tyPrim{real}{e}} t v = \lambda x. x
\\
\transdn{A_1\times A_2}t v &= \lambda p. (\transdn{A_1}t v(\pi_1 p),\transdn{A_2}t v(\pi_2 p))
\\
\transdn{A_1 + A_2} t v &= \lambda x.
\textrm{case}\ x\ \textrm{of}\ \textrm{inl}\ y.\textrm{inl}(\transdn {A_1}t v y); \textrm{inr}\ y.\textrm{inr}(\transdn {A_2} t v y)
\\ 
\transdn{A_1\to A_2} t v &= \lambda f. \lambda x. \transdn{A_2} t v (f(\transup {A_1} t v x))
\end{align*}
\begin{align*}
\transdn{\tyPrim{vec}{B,e}} t v &= \lambda x. \transdn{e}  t v (x) 
&
\transup{\tyPrim{vec}{B,e}} t v &= \lambda x. \transup{e} t v (x)
\\
\transdn{0} t v &= \lambda w.w
&
\transup{0} t v &= \lambda w.w
\\ 
\transdn{e_1+e_2} t v &= \lambda w.\transdn{e_1} t v(\transdn{e_2} t v(w))
&
\transup{e_1+e_2} t v &= \lambda w.\transup{e_1} t v(\transup{e_2} t v(w))
\\
\transdn{-e} t v &= \lambda w.\transup{e} t v (w)
&
\transup{-e} t v &= \lambda w.\transdn{e} t v (w)
\\
\transdn{t} t v &= \lambda w.w + v 
&
\transup{t} t v &= \lambda w.w - v
\\
\transdn{t'} t v &= \lambda w.w\quad(t'\neq t)
&
\transup{t'} t v &= \lambda w.w\quad(t'\neq t)
\end{align*}
To show $I(J(y))$ is contextually equivalent to $y$ is an easy
induction on $A$ using the fact that the vector $0$
is an identity for vector addition. To show $J(I(x))$ is contextually
equivalent to $x$ we must apply the abstraction theorem and use
the fact that $\tmSem{\transup A t v}$ and $\tmSem{\transdn A t v}$ 
considered as relations are simply instances of $\rsem A{}{}$.
\qed
%\[
%$\rsem A {t \mapsto v} {} = \{ (x, \tmSem{\transup A (x)}) \mid x \in
%\tySem A\}$.
%\]

\paragraph{Scalings}
For the group $\GL{1}$ of scalings we can treat
a real-valued argument to a function as a unit-of-measure with
which to scale the result. 
%The small fly in the ointment is that the
Although the argument might be zero, and this cannot be used for scaling,
%Nevertheless, 
we still can obtain the following slightly more complicated
isomorphism:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to A
& \cong &
A[1/s]\times \forall s\mathord:\SynGL{1}.A
\end{eqnarray*}
The type of $\mathrm{areaCircle}$ from the introduction is one instance:
\begin{eqnarray*}
\forall s\mathord:\SynGL{1}.\tyPrim{real}{s}\to\tyPrim{real}{s^2}\\ \quad\cong\quad \tyPrim{real}{1}\times(\forall
s\mathord:\SynGL{1}.\tyPrim{real}{s^2})\\ \quad\cong\quad
\tyPrim{real}{1}\times 1\quad\cong\quad \tyPrim{real}{1}
\end{eqnarray*}
%where %the isomorphism 
The abstraction theorem is used to prove
$\forall s\mathord:\SynGL{1}.\tyPrim{real}{s^2}\cong 1$.
% is proved using
%the abstraction theorem.
\end{example}
%\begin{enumerate}
%\item Two cases: when we have a total ``division'' operation, and when
%  we don't.
%\item Type isomorphisms, based on Smith Normal Form
% \item Non-definability results, based on the non-compositional
%   relational environments
% \item Alteration of the non-definability results in the light of
%   addition of a square root operation
% \item Mention the variation we can obtain by changing some types,
%   e.g. the type of $\mathrm{abs}$. Also, the complex numbers example.
%\end{enumerate}
